---
title: "Alcohol vs. Happiness: an Investigation into Alcoholism"
author: "Adam, Dillon, Ruben, Nick"
format: source
  html:
    self-contained: true
    code-tools: true
    toc: true
    number-sections: true
    code-fold: true
editor: source
execute:
  error: true
  echo: true
  message: false
  warning: false
---

## Setup

```{r Setup}
library(tidyverse)
library(here)
library(broom)
library(patchwork)
library(DT)
alc <- read_csv(here::here("sh_alc_pcap_li.csv"))
happy <- read_csv(here::here("hapiscore_whr.csv"))
```


## Intro

As college students, our group was especially interested regarding the implications alcohol has on the quality of life. So, we set out to investigate the existence of such association. Based on empirical data collected in [“The Association between Mental Wellbeing, Levels of Harmful Drinking, and Drinking Motivations: A Cross-Sectional Study of the UK Adult Population,”](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6069385/) which demonstrates that low-risk drinking makes for higher well-being than abstinence, harmful drinking, and hazardous drinking, we hypothesize that happiness has a positive association with alcohol consumption until a certain point at which it will being to have a negative association. This certain point will likely be determined in our analysis.



## Happiness Dataset

The “happiness score” dataset measures the national average response to a Cantril life ladder survey from 2004 to 2020. The Cantril life ladder is a well-being quantitative assessment based on an imaginary step ladder where the bottom rung is minimum life satisfaction (0) and the top rung is the best possible life satisfaction (10), explained further [here.](https://news.gallup.com/poll/122453/understanding-gallup-uses-cantril-scale.aspx.) Scores were then scaled from 0-100 to effectively communicate percentages.


## Alcohol Dataset

The “alcohol consumption per capita” dataset measures projected estimates for the quantity of alcohol consumed in people over the age of 15 from 1999 to 2017, in liters. Data were collected quinquennially, with the exception of the 2014-2017 timespan, where data were collected triennially. 


## Data Cleaning
Both datasets were then pivoted longer, and "year" columns were converted to numeric variables. We also realized there was not much justification to keeping the many years from the alcohol dataset that were entirely missing.

```{r}
# Pivoting Longer, convert years to numeric and dropping unnecessary missing alcohol columns
alc_final <- alc |> 
  select("country", "1999", "2004", "2009", "2014", "2017") |> 
  pivot_longer(cols = "1999":"2017",
               names_to = "year",
               values_to = "alcohol_consumption")|>  
  mutate(
    across("year", as.numeric))


happy_final <- happy |> 
  pivot_longer(cols = `2004`:`2020`,
               names_to = "year",
               values_to = "happiness_score")|>  
  mutate(
    across("year", as.numeric))
```


Then, we joined the datasets using an inner join so we could keep only the years both datasets had in common.

```{r}
# Inner Joining the data
final_data <- inner_join(happy_final, alc_final, by = c("year", "country"))
final_data
```
## Data Visualization

```{r}
ggplot(data = final_data, mapping = aes(x = alcohol_consumption, y = happiness_score)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  theme(axis.title.y = element_blank()) +
  labs(title = "Happiness Score By Alcohol Consumption",
       x = "Average Alcohol Consumption",
       subtitle = "Average Happiness Score",
       y= "") +
  theme_bw()
```

It appears from our linear regression model investigating the strength of the association between mean total alcohol consumption per capita in liters and mean happiness score that there is a weak positive association between these variables.

```{r}
average_year <- final_data |> 
  group_by(year) |> 
  summarize(mean_happy = mean(happiness_score, na.rm = TRUE),
            mean_alc = mean(alcohol_consumption, na.rm = TRUE))
plot1 <- ggplot(data = average_year, mapping = aes(x = year, y = mean_happy)) +
  geom_line(color = "blue") +
  labs(title = "Average Happiness Over Time (Pts)",
       x = "Year",
       y= "",
       subtitle = "Average Happiness Score") +
  theme_bw()

plot2 <- ggplot(data = average_year, mapping = aes(x = year, y = mean_alc)) +
  geom_line(color = "red") +
  labs(title = "Average Alcohol Consumption (L) \nOver Time",
       x = "Year",
       y= "",
       subtitle = "Average Alcohol Consumption") +
  theme_bw()

plot1 + plot2 
```

Based on our time series plots for mean happiness score and mean total alcohol consumption per capita in liters, both happiness and alcohol consumption have decreased over time.

## Linear Regression
```{r}
my_model <- lm(happiness_score ~ alcohol_consumption, 
               data = final_data
               )
my_model |> tidy()
```
$$
\widehat{happiness~score} = 1.001167(alcohol~consumption) + 48.580332
$$

Based on the model we fit predicting mean happiness score based on mean total alcohol consumption per capita, we can see that for every 1 liter increase in mean total alcohol consumption, happiness score increases by approximately 1.001. At 0 liters mean total alcohol consumption per capita, we predict happiness score to be approximately 48.58. 

## Model Fit

```{r}
augmented <- augment(my_model)
response_var <- var(augmented$happiness_score)
fitted_var <- var(augmented$.fitted)
residual_var <- var(augmented$.resid)
table <- data.frame(response_var, fitted_var, residual_var)
table |> datatable(class = "cell-border stripe hover", 
                   colnames = c("Variation in Response",
                                "Variation in the Fitted Values",
                                "Variation in the Residuals"),
                   rownames = FALSE,
                   options = list(paging = FALSE, 
                                  searching = FALSE, 
                                  info = FALSE))
```

```{r}
model_fit <- round((fitted_var/(response_var))*100,
                   digits = 3)
```

Based on the $R^{2}$ we calculated for the model we fit, only `r model_fit`% of variability in the response values were accounted for by the regression model. With such a low $R^{2}$ this confirms that our model is unfit for the data.