---
title: "Alcohol vs. Happiness: an Investigation into Alcoholism"
author: "Adam Del Rio, Dillon Murphy, Ruben Jimenez, Nick Patrick"
format:
  html:
    self-contained: true
    code-tools: true
    toc: true
    code-fold: true
    mainfont: "Times New Roman"
editor: source
execute:
  error: true
  echo: true
  message: false
  warning: false
---


```{r Setup}
#| include: false
library(tidyverse)
library(here)
library(broom)
library(patchwork)
library(DT)
library(kableExtra)
alc <- read_csv(here::here("sh_alc_pcap_li.csv"))
happy <- read_csv(here::here("hapiscore_whr.csv"))
```


## Introduction

As college students, our group was especially interested regarding the implications alcohol has on the quality of life. So, we set out to investigate the existence of such association. Based on empirical data collected in [“The Association between Mental Wellbeing, Levels of Harmful Drinking, and Drinking Motivations: A Cross-Sectional Study of the UK Adult Population,”](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6069385/) which demonstrates that low-risk drinking makes for higher well-being than abstinence, harmful drinking, and hazardous drinking, we hypothesize that happiness has a positive association with alcohol consumption until a certain point at which it will being to have a negative association. This certain point will likely be determined in our analysis.



The “happiness score” dataset measures the national average response to a Cantril life ladder survey from 2004 to 2020. The Cantril life ladder is a well-being quantitative assessment based on an imaginary step ladder where the bottom rung is minimum life satisfaction (0) and the top rung is the best possible life satisfaction (10), explained further [here.](https://news.gallup.com/poll/122453/understanding-gallup-uses-cantril-scale.aspx.) Scores were then scaled from 0-100 to effectively communicate percentages.



The “alcohol consumption per capita” dataset measures projected estimates for the quantity of alcohol consumed in people over the age of 15 from 1999 to 2017, in liters. Data were collected quinquennially, with the exception of the 2014-2017 timespan, where data were collected triennially. 


## Data

Both datasets were pivoted longer, and "year" columns were converted to numeric variables. We also realized there was not much justification to keeping the many years from the alcohol dataset that were entirely missing.

```{r}
# Pivoting Longer, convert years to numeric and dropping unnecessary missing alcohol columns
alc_final <- alc |> 
  select("country", "1999", "2004", "2009", "2014", "2017") |> 
  pivot_longer(cols = "1999":"2017",
               names_to = "year",
               values_to = "alcohol_consumption")|>  
  mutate(
    across("year", as.numeric))


happy_final <- happy |> 
  pivot_longer(cols = `2004`:`2020`,
               names_to = "year",
               values_to = "happiness_score")|>  
  mutate(
    across("year", as.numeric))
```


Then, we joined the datasets using an inner join so we could keep only the years both datasets had in common.

```{r}
# Inner Joining the data
final_data <- inner_join(happy_final, alc_final, by = c("year", "country"))
final_data |> 
  datatable(class = "cell-border stripe hover", 
                   colnames = c("Country",
                                "Year",
                                "Mean Happiness Score",
                                "Mean Alcohol Consumption"),
                   rownames = FALSE,
                   options = list(paging = TRUE, 
                                  searching = TRUE, 
                                  info = TRUE))
```


## Linear Regression

```{r}
# Font formatting credit: https://ggplot2.tidyverse.org/articles/faq-customising.html
ggplot(data = final_data, mapping = aes(x = alcohol_consumption, y = happiness_score)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  labs(title = "Happiness Score By Alcohol Consumption",
       x = "Average Alcohol Consumption (L)",
       subtitle = "Average Happiness Score (pts)",
       y= NULL) +
  theme_bw(base_family = "Times New Roman")
```

It appears from our linear regression model investigating the strength of the association between mean total alcohol consumption per capita in liters and mean happiness score that there is a weak positive association between these variables.

```{r}
average_year <- final_data |> 
  group_by(year) |> 
  summarize(mean_happy = mean(happiness_score, na.rm = TRUE),
            mean_alc = mean(alcohol_consumption, na.rm = TRUE))

plot1 <- ggplot(data = average_year, mapping = aes(x = year, y = mean_happy)) +
  geom_line(color = "blue") +
  labs(title = "Average Happiness Over Time (pts)",
       x = "Year",
       y= "",
       subtitle = "Average Happiness Score") +
  theme_bw(base_family = "Times New Roman")

plot2 <- ggplot(data = average_year, mapping = aes(x = year, y = mean_alc)) +
  geom_line(color = "red") +
  labs(title = "Average Alcohol Consumption \nOver Time (L)",
       x = "Year",
       y= "",
       subtitle = "Average Alcohol Consumption") +
  theme_bw(base_family = "Times New Roman")

plot1 + plot2 
```

For the plots above, we grouped the observations by year and found the mean happiness score and mean alcohol consumption per capita in liters. Based on our time series plots for mean happiness score and mean total alcohol consumption per capita in liters, both happiness and alcohol consumption have decreased over time. From 2004-2009, happiness decreased very rapidly from an average score around 64 happiness points all the way down to about 55 happiness points. Meanwhile average alcohol consumption was actually increasing very slightly. Past 2009, average happiness began to level off, and even increases after 2014. On the other hand, average alcohol consumption rapidly dropped from just under 6.35 liters per capita down to less than 6.1 liters.

## Model Fit

```{r}
my_model <- lm(happiness_score ~ alcohol_consumption, 
               data = final_data
               )
my_model |> 
  tidy() |>
  mutate(term = case_when(term == "alcohol_consumption" ~ "Alcohol Consumption",
                   term == "(Intercept)" ~ "Intercept")) |>
  kbl(col.names = c("Term",
                    "Estimate",
                    "Standard Error",
                    "Statistic",
                    "P-value"),
      caption = "Regression Output",
      digits = 4) |>
  kable_classic_2(html_font = "Times New Roman",
                  lightable_options = "striped")
```
$$
\widehat{happiness~score} = 1.001167(alcohol~consumption) + 48.580332
$$

Based on the model we fit predicting mean happiness score based on mean total alcohol consumption per capita, we can see that for every 1 liter increase in mean total alcohol consumption, happiness score increases by approximately 1.001. At 0 liters mean total alcohol consumption per capita, we predict happiness score to be approximately 48.58. 


```{r}
augmented <- augment(my_model)
response_var <- var(augmented$happiness_score) |>
  round(digits = 3)
fitted_var <- var(augmented$.fitted) |> 
  round(digits = 3)
residual_var <- var(augmented$.resid) |> 
  round(digits = 3)
table <- data.frame(response_var, fitted_var, residual_var)
table |> 
  kbl(col.names = c("Variation in Response",
                  "Variation in the Fitted Values",
                  "Variation in the Residuals"),
      caption = "Variation",
      digits = 3) |>
  kable_classic_2(html_font = "Times New Roman",
                  lightable_options = "striped")
```

```{r}
model_fit <- round((fitted_var/(response_var))*100,
                   digits = 3)
```

By dividing the amount of variation in the fitted values (values from our predictive model) by the total variation in the observed responses, we obtain an $R^{2}$ statistic representing the amount of variation explained by our model. In other words, this statistic explains how well our model can predict the pattern in the data. Based on the $R^{2}$ we calculated for the model we fit, only `r model_fit`% of variability in the response values were accounted for by the regression model. With such a low $R^{2}$ this confirms that our model is unfit for the data.


### Simulation

```{r}
noise <- function(x, mean = 0, sd){
  x + rnorm(length(x), 
            mean, 
            sd)
}
```


```{r}
preds <- predict(my_model) 

sd <- sigma(my_model)

final_preds <- tibble(happiness_score = noise(preds, 
                                           sd = sd)
                   )

final_preds
```


```{r}
ggplot(data = final_data, mapping = aes(x = alcohol_consumption, y = happiness_score)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  labs(title = "Happiness Score By Alcohol Consumption",
       x = "Average Alcohol Consumption (L)",
       subtitle = "Average Happiness Score (pts)",
       y= NULL) +
  theme_bw(base_family = "Times New Roman")
```

